---
title: \vspace{-2cm} Logistic Regression Modeling of the Growth Limit of Alicyclobacillus Acidoterrestris CRA7152 in Apple Juice
author: "Loftsson, Valentin Oliver — valentin.loftsson@epfl.ch"
geometry: margin=2.5cm
fontsize: 12pt
output:
  pdf_document:
    latex_engine: xelatex
  html_document:
    df_print: paged
  word_document: default
mainfont: Calibri Light
bibliography: bibliography.bib
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[L]{\textit{\leftmark}}
- \fancyhead[R]{Loftsson}
- \usepackage{titlesec}
- \titlespacing{\title}{0pt}{\parskip}{-\parskip}
- \usepackage{hyperref}
- \hypersetup{colorlinks=true, urlcolor=black}
- \usepackage{bm}
- \usepackage{amsmath}
- \usepackage[font={small, color=blue}]{caption}
- \usepackage{subcaption}
---

```{r setup, include=F}
knitr::opts_chunk$set(echo=F, eval=F, warning=F, message=F)
# library(dplyr)
# library(tidyr)
# library(caret)
library(stats)
# library(knitr)
library(kableExtra)
library(plotly)
```

```{r load_data}
features <- c("ph", "nisin", "temperature", "brix")
target <- "growth"
colnames <- c(features, c(target))
apple_juice <- read.table(
  "apple_juice.dat",
  header = FALSE,
  col.names = colnames,
)
factors <- c("No Growth", "Growth")
apple_juice$growth <- factor(
  apple_juice$growth,
  labels = factors
)

attach(apple_juice)
```

```{r}
logreg <- function(rhs) {
  formula <- as.formula(paste("growth ~ ", rhs))
  glm(formula, data=apple_juice, family=binomial())
}
```

```{r}
kable.var <- function(dt, caption, file) {
  dt %>%
    kable(caption = caption) %>%
    kable_styling(full_width = F) %>%
    row_spec(5, bold = T)
}

table.ph <- addmargins(table(ph, growth))
table.nisin <- addmargins(table(nisin, growth))
table.temperature <- addmargins(table(temperature, growth))
table.brix <- addmargins(table(brix, growth))

kable.var(table.ph, "Table 1: pH")
kable.var(table.nisin, "Table 2: Nisin concentration (IU/ml)")
kable.var(table.temperature, "Table 3: Temperature (°C)")
kable.var(table.brix, "Table 4: Soluble solids conc. (°Brix)")
```

```{r}
# All main and interaction effects
rhs.features <- paste("(", paste(features, collapse =" + "), ")^2")

# Model with all main and interaction effects
glm.full <- logreg(rhs.features)

# LRT
drop1.res <- drop1(glm.full, test="LRT")

# Final model
terms.dropped <- " -ph:nisin -ph:temperature -nisin:temperature"
glm.final <- logreg(paste(rhs.features, terms.dropped))
glm.final.summary <- summary.glm(glm.final)
```

```{r}
drop1.lrt <- drop1.res[4] %>% round(2)
drop1.lrt["P value"] <- drop1.res$`Pr(>Chi)` %>% signif(4)
drop1.lrt <- drop1.lrt[2:7,]
caption <- "Table 5: LRT test on interaction effects"
drop1.lrt %>%
  kable(caption = caption, format = "html") %>%
  kable_styling(full_width = F, bootstrap_options = "basic", font_size = 20) %>%
  column_spec(column = 1, width = "1in") %>%
  column_spec(column = 2:3, width = "1.5in") %>%
  save_kable("table5.png")
```

```{r}
coefs <- round(glm.final.summary$coefficients[,c(1, 4)], digits = 4)
caption <- "Table 6: Coefficients of the final model for A. Acidoterrestris growth in apple juice"
coefs
coefs %>%
    kable(caption = caption, col.names = c("Coefficient", "P value"), format = "html") %>%
    kable_styling(full_width = F, bootstrap_options = "basic", font_size = 20) %>%
    save_kable("table6.png")
```

```{r}
confusion.matrix <- function(m) {
  # type=response gives predicted probabilities
  model_prob <- predict(m, apple_juice, type="response")
  model_pred <- factor(1*(model_prob > .5), labels=factors)
  model_real <- factor(1*(growth == "Growth"), labels=factors)
  table(model_real, model_pred)
}

caption <- "Table 7: Confusion matrix summarizing prediction results from applying the final model on the data samples"
cm.table <- confusion.matrix(glm.final)
cm.table %>%
  kable(caption = caption, label = caption) %>%
  kable_styling(full_width = F, bootstrap_options = "basic", font_size = 20) %>%
  column_spec(column = 1, bold = TRUE) %>%
  add_header_above(c("Real response" = 1, "Predicted response" = 2)) %>%
  save_kable("table7.png")
```

```{r}
prob.grid.temp.brix <- function(ph.const, nisin.const) {
  dimrow <- max(temperature)
  dimcol <- max(brix)
  grid <- array(dim = c(dimrow, dimcol))
  for (row in 1:dimrow) {
    for (col in 1:dimcol) {
      newdata <- data.frame(ph=ph.const, nisin=nisin.const, temperature=row, brix=col)
      grid[row, col] = predict(glm.final, newdata, type="response")
    }
  }
  
  grid
}

prob.grid.nisin.brix <- function(ph.const, temp.const) {
  dimrow <- max(nisin)
  dimcol <- max(brix)
  grid <- array(dim = c(dimrow, dimcol))
  for (row in 1:dimrow) {
    for (col in 1:dimcol) {
      newdata <- data.frame(ph=ph.const, nisin=row, temperature=temp.const, brix=col)
      grid[row, col] = predict(glm.final, newdata, type="response")
    }
  }
  
  grid
}
```

```{r}
plot.surface <- function(filename, data, title, xlabel, ylabel, minx, miny) {
  x <- minx:ncol(data)
  y <- miny:nrow(data)
  z <- data[y, x]
  default.font = list(size = 14)
  fig <- plot_ly(x=x, y=y, z=z, type="surface", colors="YlOrRd")
  fig <- fig %>% layout(
    font = default.font,
    title = list(
      text = title,
      font = list(size = 17),
      y = 0.95
    ),
    scene = list(
      xaxis = list(
        title = list(
          text = xlabel,
          font = default.font
        )
      ),
      yaxis = list(
        title = list(
          text = ylabel,
          font = default.font
        )
      ),
      zaxis = list(
        title = list(
          text = "growth probability",
          font = default.font
        )
      )
    ),
    height = 650,
    margin = list(t = 0)
  )
  api_create(fig, filename = filename)
  fig
}
```

```{r}
xlabel <- "soluble solids conc. (°Brix)"
ylabel <- "temperature (°C)"

g <- prob.grid.temp.brix(ph.const = 3.7, nisin.const = 0)

plot.surface(
  "loftsson-apple-juice-cra7152-growth-figure-1a",
  g,
  title = "Figure 1.a: Growth probability of CRA7152 in\napple juice with <b>nisin 0 IU/ml and pH = 3.7</b>",
  xlabel,
  ylabel,
  minx = min(brix),
  miny = min(temperature)
)
```

```{r}
g <- prob.grid.temp.brix(ph.const = 4.5, nisin.const = 0)

plot.surface(
  "loftsson-apple-juice-cra7152-growth-figure-1b",
  g,
  title = "Figure 1.b: Growth probability of CRA7152 in\napple juice with <b>nisin 0 IU/ml and pH = 4.5</b>",
  xlabel,
  ylabel,
  minx = min(brix),
  miny = min(temperature)
)
```

```{r}
xlabel <- "soluble solids conc. (°Brix)"
ylabel <- "nisin conc. (UI/ml)"

g <- prob.grid.nisin.brix(ph.const = 4.0, temp.const = 45)

plot.surface(
  "loftsson-apple-juice-cra7152-growth-figure-2a",
  g,
  title = "Figure 2.a: Growth probability of CRA7152 in\napple juice with <b>pH = 4.5 and temp. = 45°C</b>",
  xlabel,
  ylabel,
  minx = min(brix),
  miny = min(nisin)
)
```

```{r}
g <- prob.grid.nisin.brix(ph.const = 4.0, temp.const = 30)

plot.surface(
  "loftsson-apple-juice-cra7152-growth-figure-2b",
  g,
  title = "Figure 2.b: Growth probability of CRA7152 in\napple juice with <b>pH = 4.5 and temp. = 30°C</b>",
  xlabel,
  ylabel,
  minx = min(brix),
  miny = min(nisin)
)
```

# Introduction
Microbiological contamination in processed acid fruit juices has been the subject of microbiological interest and research in the past few decades. Controlling such contamination is an important factor in the preservation of packed fruit juice products. To minimize contamination risk, it is essential to consider changes in the microbiology of the product over the entire production cycle, from cultivation and harvesting to preparation, distribution, and storage.

Peña *et al.* [-@pena] studied the contamination in apple juice caused by an acidophilic sporeformer bacterium, *Alicyclobacillus acidoterrestris* CRA7152, and developed a model to predict its growth probability. The bacterium causes "off-flavor" in fruit juices from certain chemical compounds. Several studies have shown that *Alicyclobacillus acidoterrestris* grows at temperatures between 25 to 60°C [@previdi] with pH varying from 3.0 to 5.5 [@eguchi]. The use of nisin, a chemical that is used as a food preservative, has proven effective as a minimum inhibitory concentration method to control the germination of the bacterium's spores. However, the effectiveness of nisin depends also on the temperature, pH, and soluble solids concentration.

Evaluating the bacterium's growth/nongrowth interface becomes cumbersome when multiple factors are involved. Logistic regression is a powerful tool for probabilistic microbial modeling, given that sufficient information is available about the product characteristics and conditions. It has been used frequently in similar studies. With logistic regression, it is possible to describe the growth probability and establish critical limits for the predictive variables that determine the condition of the apple juice.

In this paper, we reproduce the statistical analyses performed by Peña *et al.* on the original dataset. The main purpose of the original study is to "evaluate the growth responses of *Alicyclobacillus acidoterrestris* in apple juice with different pH values, concentrations of soluble solids and nisin subjected to different incubation temperatures". We begin with an exploratory data analysis and then perform a logistic regression analysis to determine and evaluate an optimal predictive model.

# Exploratory Data Analysis
The dataset includes 74 samples of apple juice assays including a binary growth response (0=No Growth, 1=Growth) evaluated after 16 days of incubation. The dataset consists of two collections of 37 assays, tested in duplicate. The assays were carefully created by the researchers to ensure a balanced dataset with a varied combination of factors that determine the juice condition.

The effect of four factors was studied as summarized in \hyperref[table1:ph]{Tables 1-4}. The frequency table of the growth response reveals that we have a semi-balanced dataset, with several more negative examples and growth occurring in 26 out of 74 assays ($\sim$ 35%). For each factor, four discrete values were selected for testing after careful experimentation and analysis, and considering accumulated knowledge from previous studies.  The tables reveal some interesting characteristics about individual factors independent of interaction effects: Growth was completely inhibited at pH 3.5 (\hyperref[table1:ph]{Table 1}) which is the minimum pH of the juice. It was also fully inhibited at the maximum nisin concentration 70 IU/ml (\hyperref[table2:nisin]{Table 2}) and at the maximum soluble solids concentration of 19 °Brix (\hyperref[table4:brix]{Table 4}), which marks the concentration level above which no growth of *A. acidoterrestris* occurs [@splittstoesser].

\begin{figure}
\centering
\begin{minipage}{.5\textwidth}
  \centering
  \caption*{}
  \includegraphics[width=\linewidth]{ph}
  \label{table1:ph}
\end{minipage}%
\begin{minipage}{.5\textwidth}
  \centering
  \caption*{}
  \includegraphics[width=\linewidth]{nisin}
  \label{table2:nisin}
\end{minipage}
\end{figure}

\begin{figure}
\centering
\begin{minipage}{.5\textwidth}
  \centering
  \caption*{}
  \includegraphics[width=\linewidth]{temperature}
  \label{table3:temperature}
\end{minipage}%
\begin{minipage}{.5\textwidth}
  \centering
  \caption*{}
  \includegraphics[width=\linewidth]{brix}
  \label{table4:brix}
\end{minipage}
\end{figure}

# Model Fitting and Selection
Logistic regression is used as a method for model fitting as it is suitable for a binary response. Here, we use the logit function as the link transformation for the generalized model:

\begin{equation}
logit(p) = \ln{\frac{p}{1-p}} = \beta_0 + \beta_1x_1 + \beta_2x_2 + \dots + \beta_kx_k
\end{equation}

Essentially, this means that the linear transformation outputs the logit transformed probability of a positive response ($Y=Growth$). To obtain the probability, $p$, we must apply the inverse logit function:

\begin{equation}
p = \frac{\exp(\beta_0 + \beta_1x_1 + \beta_2x_2 + \dots + \beta_kx_k)}{1 + \exp(\beta_0 + \beta_1x_1 + \beta_2x_2 + \dots + \beta_kx_k)}
\end{equation}

The parameters $\beta_k$ are estimated by maximum likelihood.

To capture important characteristics of relationships between the explanatory variables, we start with a model that includes all main and interaction effects. Note that `^2` denotes the inclusion of all interaction effects:

```r
growth ~ (ph + nisin + temperature + brix)^2
```

A likelihood-ratio test (LRT) was used to perform backward variable selection. The LRT compares two models where one is a sub-model of the other, by computing the LR test statistic, denoted here as $\lambda$:

\begin{equation}
\lambda = -2\ln{\frac{L(m_1)}{L(m_2)}} = 2(\ln{L(m_2)} - \ln{L(m_1)})
\end{equation}

where $m_1$ corresponds to the more restrictive model and $m_2$ is the less restrictive model. $L(*)$ denotes the likelihood of the data samples given a model. The statistic is Chi-squared distributed.

The LRT essentially tests the following null and alternative hypotheses:

* **H**: $\beta_k = 0$ for all variables $k$ that differ between the models
* **A**: $\beta_k \neq 0$

\begin{figure}
\centering
\includegraphics{table5}
\caption*{}
\label{table5}
\end{figure}

For each factor, we compare the full model described above against the sub-model with that factor removed. \hyperref[table5]{Table 5} summarizes the results. Clearly, the terms `ph:nisin`, `ph:temperature`, and `nisin:temperature` are not significant as their P-values indicate (P > 0.05). As a result, we do not reject the null hypothesis that the coefficients for these terms are zero and they are thus dropped from the model.


# Final Model
The remaining terms of the final model are all the main effects and interactions of `brix` with each variable. The coefficients and corresponding P-values are presented in \hyperref[table6]{Table 6}. Below is the model in mathematical terms:
\begin{multline}
logit(p(Growth)) = -260.7453 \\
+ 43.0019\cdot\pmb{pH} - 1.9031\cdot\pmb{nisin} + 3.1852\cdot\pmb{temp} + 13.4122\cdot\pmb{brix} \\
- 2.2149\cdot\pmb{pH}\cdot\pmb{brix} + 0.1002\cdot\pmb{nisin}\cdot\pmb{brix} - 0.1688\cdot\pmb{temp}\cdot\pmb{brix}
\end{multline}

$~$

\begin{figure}
\centering
\includegraphics{table6}
\caption*{}
\label{table6}
\end{figure}

# Model Evaluation
The model concordance or classification accuracy was used to assess the model's performance. The resulting confusion matrix is presented in \hyperref[table7]{Table 7}. The overall accuracy is 97.3%, meaning only 2.7% or 2 samples out of 74 were incorrectly classified. The final model and accuracy results match the findings of Peña *et al.* [-@pena]. 

\begin{figure}
\centering
\includegraphics{table7}
\caption*{}
\label{table7}
\end{figure}

Peña *et al.* [-@pena] confirmed the validity of the model through experiments on new samples. These experiments testified to the usefulness of nisin in preventing the growth of the bacterium and thereby increasing the shelf life of apple juice, especially under conditions of high temperature.

We visualize the behavior of the final model on surface graphs (\hyperref[fig1a]{Figures 1-2}). The graphs show the predicted growth probability of *A. acidoterrestris* in apple juice as a function of two main effects. The graphs can be explored interactively online by clicking the links above the figure captions.

\hyperref[fig1a]{Figures 1.a and 1.b} show the probability as a function of soluble solids concentration and temperature, given 0 IU/ml of nisin and pH 3.7 and 4.5, respectively. For pH 4.5 (\hyperref[fig1b]{Figure 1.b}), growth was inhibited only in high concentrations of soluble solids (Brix). Interestingly, when pH drops down to 3.7 (\hyperref[fig1a]{Figure 1.a}) no growth occurs for any value of Brix up to a temperature of 30°C. Above 30°C, the probability is close to zero only above 18°Brix.

\hyperref[fig2a]{Figures 2.a and 2.b} show the probability as a function of soluble solids concentration and nisin concentration, given pH 4.0 and temperature 45°C and 30°C, respectively. For 45°C (\hyperref[fig2a]{Figure 2.a}) growth can be inhibited by adding 40-70 IU/ml of nisin. When the temperature is decreased to 30°C (\hyperref[fig2b]{Figure 2.b}) we observe that the model predicts zero or small probability of growth for any Brix level and low nisin levels. At zero nisin, the probability decreases as the Brix level increases.

The graphs show the effect and interactions of the variables and reveal critical values that are important to establish minimum inhibitory limits. Such limits are necessary for minimizing the use of inhibitory chemicals in the product and depend on the pH and temperature levels. Peña *et al.* compute and discuss critical boundaries in more detail.


\begin{figure}
\centering
\begin{minipage}{.5\textwidth}
  \centering
  \caption*{\hypersetup{urlcolor=blue}\href{https://chart-studio.plotly.com/~valentinoli/7/}{View interactive plot online}}
  \includegraphics[width=\linewidth]{fig1a}
  \label{fig1a}
\end{minipage}%
\begin{minipage}{.5\textwidth}
  \centering
  \caption*{\hypersetup{urlcolor=blue}\href{https://chart-studio.plotly.com/~valentinoli/9/}{View interactive plot online}}
  \includegraphics[width=\linewidth]{fig1b}
  \label{fig1b}
\end{minipage}
\end{figure}

\begin{figure}
\centering
\begin{minipage}{.5\textwidth}
  \centering
  \caption*{\hypersetup{urlcolor=blue}\href{https://chart-studio.plotly.com/~valentinoli/11/}{View interactive plot online}}
  \includegraphics[width=\linewidth]{fig2a}
  \label{fig2a}
\end{minipage}%
\begin{minipage}{.5\textwidth}
  \centering
  \caption*{\hypersetup{urlcolor=blue}\href{https://chart-studio.plotly.com/~valentinoli/13/}{View interactive plot online}}
  \includegraphics[width=\linewidth]{fig2b}
  \label{fig2b}
\end{minipage}
\end{figure}


# Conclusion
In this paper, we set out to reproduce the work of Peña *et al.* [-@pena] to model the growth probability of *Alicyclobacillus acidoterrestris* in apple juice depending on pH, incubation temperature, soluble solids, and nisin concentration. We successfully fit the data using logistic regression and found an optimal model using stepwise model selection. The results match the findings of the original study. Since the final model includes both main and interaction terms, the model coefficients are difficult to interpret. Nevertheless, the model's prediction accuracy has been verified through accuracy testing, visualization, and experimentation. The model provides a beneficial tool for apple juice production since it can be used to minimize the risk of bacterial spoilage in apple juice.

# Software
The reproduction was performed in the `R` programming language [@r]. Tables were created with kableExtra [@kableExtra]. Interactive figures were created with Plotly [@plotly].

# References
